\pkg{cubble}: An \proglang{R} Package for Organizing and Wrangling Multivariate Spatio-temporal Data

### 2023-07-02 cubble.Rmd
change cb_nested -> cb_spatial,cb_long -> cb_temporal, and alike

In a spatial cubble object, spatial variables are organised as columns and temporal variables are nested within a specialised `ts` column. For example, the spatial cubble object, \code{cb_nested} printed below, contains weather records of three airport stations from the Global Historical Climatology Network Daily (GHCND) database [@menne2012overview]. In this case, the spatial cubble is convenient for wrangling the spatial variables:

```{r}
cb_nested <- climate_mel
```

```{r echo = TRUE}
cb_nested
```

In a temporal cubble, temporal variables are expanded in the long form and spatial variables are stored as a data attribute. The temporal cubble object, \code{cb_long}, contains the same spatio-temporal data as the spatial cubble object, \code{cb_nested}, but in a structure that is easier for temporal analysis:

```{r}
cb_long <- face_temporal(climate_mel)
```

```{r echo = TRUE}
cb_long
```

@@

```{r echo = TRUE}
spatial(cb_long)
```

@@

The pair of cubble verbs, \code{face_temporal()} and \code{face_spatial()}, pivots the cubble object between its two forms or faces, as illustrated in Figure \ref{fig:face}. The code applies \code{face_temporal()} on the spatial cubble (\code{cb_nested}) introduced in Section \@ref(object) to get a temporal cubble:

```{r cubble-fun, echo = TRUE}
face_temporal(cb_nested)
```

@@

```{r cubble-fun2, echo = TRUE}
identical(face_spatial(face_temporal(cb_nested)), cb_nested)
identical(face_temporal(face_spatial(cb_long)), cb_long)
```

@@

```{r echo = TRUE}
ts_nested <- make_cubble(
  spatial = stations, temporal = meteo_ts, coords = c(long, lat))
(ts_long <- face_temporal(ts_nested))
class(ts_long)
```

@@

```{r echo = TRUE}
ts_long |> has_gaps()
```

The temporal component of a created temporal cubble can be turned to a \code{tsibble} object using \code{make_temporal_tsibble()}. See the code example below using the \code{cb_long} object, created in Section \@ref(create):

```{r echo = TRUE}
cb_long |> make_temporal_tsibble() 
```

@@

```{r echo = TRUE}
(sf_nested <- make_cubble(
  spatial = stations_sf, temporal = meteo, 
  key = id, index = date))
class(sf_nested)
```

@@

```{r echo =TRUE, message=FALSE}
sf_nested |> sf::st_transform(crs = "EPSG:3857")
```

@@

```{r echo = TRUE}
cb_nested |> make_spatial_sf() 
```

@@

````
nested <- clean |> SharedData$new(~id, group = "cubble")

long <- clean |> 
  face_temporal() |> 
  SharedData$new(~id, group = "cubble")

p1 <- nested |> ggplot() + ...
p2 <- long |> ggplot() + ...
crosstalk::bscols(plotly::ggplotly(p1), plotly::ggplotly(p2), ...)
````

>>>>

In a spatial cubble object, spatial variables are organised as columns and temporal variables are nested within a specialised `ts` column. For example, the spatial cubble object, \code{cb_spatial} printed below, contains weather records of three airport stations from the Global Historical Climatology Network Daily (GHCND) database [@menne2012overview]. In this case, the spatial cubble is convenient for wrangling the spatial variables:

```{r}
cb_spatial <- climate_mel
```

```{r echo = TRUE}
cb_spatial
```

In a temporal cubble, temporal variables are expanded in the long form and spatial variables are stored as a data attribute. The temporal cubble object, \code{cb_temporal}, contains the same spatio-temporal data as the spatial cubble object, \code{cb_spatial}, but in a structure that is easier for temporal analysis:

```{r}
cb_temporal <- face_temporal(climate_mel)
```

```{r echo = TRUE}
cb_temporal
```

@@

```{r echo = TRUE}
spatial(cb_temporal)
```

@@

The pair of cubble verbs, \code{face_temporal()} and \code{face_spatial()}, pivots the cubble object between its two forms or faces, as illustrated in Figure \ref{fig:face}. The code applies \code{face_temporal()} on the spatial cubble (\code{cb_spatial}) introduced in Section \@ref(object) to get a temporal cubble:

```{r cubble-fun, echo = TRUE}
face_temporal(cb_spatial)
```

@@

```{r cubble-fun2, echo = TRUE}
identical(face_spatial(face_temporal(cb_spatial)), cb_spatial)
identical(face_temporal(face_spatial(cb_temporal)), cb_temporal)
```

@@

```{r echo = TRUE}
ts_spatial <- make_cubble(
  spatial = stations, temporal = meteo_ts, coords = c(long, lat))
(ts_temporal <- face_temporal(ts_spatial))
class(ts_temporal)
```

@@

```{r echo = TRUE}
ts_temporal |> has_gaps()
```

The temporal component of a created temporal cubble can be turned to a \code{tsibble} object using \code{make_temporal_tsibble()}. See the code example below using the \code{cb_temporal} object, created in Section \@ref(create):

```{r echo = TRUE}
cb_temporal |> make_temporal_tsibble() 
```

@@

```{r echo = TRUE}
(sf_spatial <- make_cubble(
  spatial = stations_sf, temporal = meteo, 
  key = id, index = date))
class(sf_spatial)
```

@@

```{r echo =TRUE, message=FALSE}
sf_spatial |> sf::st_transform(crs = "EPSG:3857")
```

@@

```{r echo = TRUE}
cb_spatial |> make_spatial_sf() 
```

@@

````
sd_spatial <- clean |> SharedData$new(~id, group = "cubble")

sd_temporal <- clean |> 
  face_temporal() |> 
  SharedData$new(~id, group = "cubble")

p1 <- sd_spatial |> ggplot() + ...
p2 <- sd_temporal |> ggplot() + ...
crosstalk::bscols(plotly::ggplotly(p1), plotly::ggplotly(p2), ...)
````
